---
title: "Body mass index: Continuous data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Body mass index: Continuous data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

The `mglm4twin` package implements multivariate generalized
linear models for twin and family data as proposed by Bonat and Hjelmborg (2020). The core fit function `mglm4twin` is employed for fitting a set of models. In this introductory vignette we restrict ourselves to model 
continuous outcomes. We present a set of special specifications of the standard ACE for continuous (Gaussian) trait in the context of twin data. This vignette reproduces the data analysis presented in the Section 5.2 Dataset 2: Continuous data of Bonat and Hjelmborg (2020).

## Install and loading the mglm4twin package

To install the stable version of [`mglm4twin`][], use
`devtools::install_git()`. For more information, visit [mglm4twin/README].

```{r, eval=FALSE}
library(devtools)
install_git("wbonat/mglm4twin")
```

```{r, eval=TRUE, error=FALSE, message=FALSE, warning=FALSE}
library(mglm4twin)
packageVersion("mglm4twin")
```

## Data set

This example regards a fairly common continuous trait analysis. 
The dataset is part of the `twinbmi`dataset from the `mets`package.
In this vignette, we restrict our attention to a subset of the original data set where we have paired observation for both twin. The resulting dataset consists of 4271 (2788 DZ and 1483 MZ) twin-pairs.
The analysis goal is to investigate the genetic and common environment
influences on the body mass index (BMI). 

```{r}
data(bmi)
head(bmi)
```

## Model specification

The model specification is done in three steps:

   1. Specifying the linear predictor (mean model).
   2. Specifying the matrix linear predictor (twin dependence struture).
   3. Specifying link and variance function suitable to be response variable type.

### Specyfing the linear predictor

The linear predictor was composed of the interactive effect between zygosity (`Group`) (DZ and MZ) and twin pair (`Twin_pair`) code (Twin 1 and Twin 2). Thus, the linear predictor is given by

```{r}
linear_pred <- bmi ~ Group*Twin_pair
```

### Specyfing the matrix linear predictor

The matrix linear predictor describes the twin dependence structure. The `mglm4twin` package provides a set of pre-specified covariance structures.
The matrix linear predictor is a linear combination of known matrices. 
This specification allows for an easy and flexible specification of the covariance structure including the possibility to model each of the dispersion components as a linear function of known covariates.
In this example, we model the dispersion components for the ACE model as a function of age and gender. Let's start specifying the standard ACE and AE models.

```{r}
ACE = mt_twin(N_DZ = 5576/2, N_MZ = 2966/2, n_resp = 1, model = "ACE")
AE = mt_twin(N_DZ = 5576/2, N_MZ = 2966/2, n_resp = 1, model = "AE")
```

For specifying the linear models for the dispersion components the `mglm4twin`package provides a an interface similar to the `glm()` using formulas. Let's standardized the covariate age to avoid numerical issues and then specify the formula for each of the ACE model's components.

```{r}
bmi$age <- (bmi$age - mean(bmi$age))/sd(bmi$age)
list_form <- list(formE = ~ age + gender, formA = ~ age + gender,
                  formC = ~ age + gender)
```

By using this set of formula we can specify the matrix linear predictor for the ACE model. Note that, now we are using the argument `formula` where we pass a list of formulas.

```{r}
ACE_reg = mt_twin(N_DZ = 5576/2, N_MZ = 2966/2, n_resp = 1,
                  model = "ACE", formula = list_form, data = bmi)
```

We can restrict the more general ACE model to its special cases as for example the AE model where the components A and E are modelled as a function of `age` and `gender`.

```{r}
list_form2 <- list(formE = ~ age + gender, formA = ~ age + gender)
AE_reg = mt_twin(N_DZ = 5576/2, N_MZ = 2966/2, n_resp = 1,
                 model = "AE", formula = list_form2, data = bmi)
```

### Specyfing link and variance functions

The specificantion of the link and variance functions depends on the response variable type. In the case of the continuous and symmetric data 
the standard choice is the `identity` function.
For the variance function the `constant` represents the standard Gaussian variance function, which is suitable for continuous and symmetric outcomes. In this example, we follow Bonat and Hjelmborg (2020) and use the `identity` link function and the `constant` variance function. It is important to note that these choices are the default approach in the `mglm4twin()`function. 

```{r}
link = "identity"
variance = "constant"
```

## Model fitting

The next step is fitting the specified models.

```{r}
## Standard ACE model
fit_ACE <- mglm4twin(linear_pred = c(linear_pred), 
                     matrix_pred = ACE, data = bmi)

## Standard AE model
fit_AE <- mglm4twin(linear_pred = c(linear_pred), 
                    matrix_pred = AE, data = bmi)

## ACE regression on the dispersion
fit_ACE_reg <- mglm4twin(linear_pred = c(linear_pred), 
                         matrix_pred = ACE_reg, data = bmi)

## AE regression on the dispersion
fit_AE_reg <- mglm4twin(linear_pred = c(linear_pred), 
                        matrix_pred = AE_reg, data = bmi)
```

## Model output

In this example, there are a lot of dispersion components and model selection is required to simplify the model by dropping non-significant components. The `mglm4twin` package offers a set of functions to assist in this process. The `mt_anova_mglm()`function performs an ANOVA-type test for the dispersion components in each of the ACE model components.

```{r}
mt_anova_mglm(fit_ACE_reg, formula = list_form, data = bmi)
```

The results show that estimates involved in the C component are non-significant. Thus, we simplify the model to the AE. Checking the
significance of the AE model's components.

```{r}
mt_anova_mglm(fit_AE_reg, formula = list_form2, data = bmi)
```

The results show that in this case all estimates of the E and A are jointly significant. We can further explore the significance of the components by looking at each one. When using `type = robust2` it can take long that's why we do not run this code while building the package, but the user can run without problems.

```{r, eval = FALSE}
aux_summary(fit_AE_reg, formula = list_form2, 
            type = "robust2", id = bmi$Twin, data = bmi)
```

The results show that the A dispersion component associate to the age is non-significant. Thus, we can specify an AE simplified model by dropping this component.

```{r}
list_form3 <- list(formE = ~ age + gender, formA = ~ gender)
AE_reg1 = mt_twin(N_DZ = 5576/2, N_MZ = 2966/2, n_resp = 1,
                  model = "AE", formula = list_form3, data = bmi)
fit_AE_reg1 <- mglm4twin(linear_pred = c(linear_pred), matrix_pred = AE_reg1,
                         data = bmi)
```

#### Standard output

For complex model as the ones with covariates describing the dispersion structure the `mglm4twin` package does not offer standard model's output like `summary()`.
Thus, the users have to extract the componets and build the output themselves.

## Reproducing Table 2 of Bonat and Hjelmborg (2020)

```{r, echo = FALSE, results = "hide"}
Parameters <- c("E0","E_age","E_gender", "A0", "A_age", "A_gender",
                "C0", "C_age", "C_gender", "plogLik(df)", "pAIC")

## ACE regression model
tt = aux_summary(fit_ACE_reg, formula = list_form, type = "otimist", 
                 id = bmi$Twin, data = bmi)
std_ACE_reg <- tt$Std..Error
ACE_reg <- fit_ACE_reg$Covariance

## AE regression model
AE_reg <- c(fit_AE_reg$Covariance, rep(NA, 3))
tt = aux_summary(fit_AE_reg, formula = list_form2, type = "otimist", data = bmi)
std_AE_reg <- tt$Std..Error
std_AE_reg <- c(std_AE_reg, rep(NA, 3))

## AE simplified 
AE_reg1 <- c(fit_AE_reg1$Covariance[1:3], fit_AE_reg1$Covariance[4], NA,
             fit_AE_reg1$Covariance[5], rep(NA, 3))
tt = aux_summary(fit_AE_reg1, formula = list_form3, type = "otimist", data = bmi)
std_AE_reg1 <- tt$Std..Error
std_AE_reg1 <- c(std_AE_reg1[1:3], std_AE_reg1[4], NA, std_AE_reg1[5], rep(NA, 3))

## Table 2
Table2 <- data.frame(round(cbind(ACE_reg, std_ACE_reg,AE_reg, std_AE_reg, AE_reg1, std_AE_reg1),2))
LL <- c(plogLik(fit_ACE_reg, verbose = FALSE)$plogLik, 
        plogLik(fit_ACE_reg, verbose = FALSE)$df,
        plogLik(fit_AE_reg, verbose = FALSE)$plogLik, 
        plogLik(fit_AE_reg, verbose = FALSE)$df,
        plogLik(fit_AE_reg1, verbose = FALSE)$plogLik, 
        plogLik(fit_AE_reg1, verbose = FALSE)$df)

pAIC <- c(pAIC(fit_ACE_reg, verbose = FALSE)$pAIC, NA, 
          pAIC(fit_AE_reg, verbose = FALSE)$pAIC, NA, 
          pAIC(fit_AE_reg1, verbose = FALSE)$pAIC, NA)
Table2 <- rbind(Table2, LL, pAIC)
Table2 <- cbind(Parameters, Table2)
```

```{r}
Table2
```

## Reproducing Figure 5 of Bonat and Hjelmborg (2020)

```{r, echo = FALSE, out.width='99%', message=FALSE, fig.width=5, fig.height=5, fig.align='center', fig.cap = 'Environment, additive genetic and heritability index as a function of standardized age.'}
phiE = fit_AE_reg1$Covariance[1:3]
phiA = fit_AE_reg1$Covariance[4:5]
data_new <- data.frame("age" = rep(seq(from = -1.539967, to = 2.098272, length = 30), 2),
                       "gender" = rep(c("female", "male"), each = 30))
X_E <- model.matrix(~ age + gender, data = data_new)
X_A <- model.matrix(~ gender, data = data_new)
Point <- fit_AE_reg1$Covariance
VCOV <- fit_AE_reg1$vcov[5:9, 5:9]

## Simulation (computin standard  errors and confidence interval)
require(mvtnorm)
require(lattice)
require(latticeExtra)
source("https://raw.githubusercontent.com/walmes/wzRfun/master/R/panel.cbH.R")
tt = rmvnorm(n = 5000, mean = Point, sigma = as.matrix(VCOV))

# Computing measures of interest
tauE <- matrix(NA, nrow = dim(X_E)[1], ncol = 5000)
tauA <- matrix(NA, nrow = dim(X_E)[1], ncol = 5000)
h <- matrix(NA, nrow = dim(X_E)[1], ncol = 5000)

for(i in 1:5000) {
  tauE[,i] <- X_E%*%tt[i,1:3]
  tauA[,i] <- X_A%*%tt[i,4:5]
  h[,i] <- tauA[,i]/(tauE[,i] + tauA[,i])
}

# Environment
data_new$tauE_est <- apply(tauE, 1, mean)
tauE_IC <- apply(tauE, 1, quantile, c(0.025, 0.975))
data_new$tauE_IcMin <- tauE_IC[1,]
data_new$tauE_IcMax <- tauE_IC[2,]

# Genetic
data_new$tauA_est <- apply(tauA, 1, mean)
tauA_IC <- apply(tauA, 1, quantile, c(0.025, 0.975))
data_new$tauA_IcMin <- tauA_IC[1,]
data_new$tauA_IcMax <- tauA_IC[2,]

# Heritability
conf_h = data.frame(t(apply(h, 1, quantile, c(0.025, 0.975))))
data_new$IcMin_h <- conf_h[,1]
data_new$IcMax_h <- conf_h[,2]
data_new$h <- rowMeans(h)
levels(data_new$gender) <- c("Female", "Male")

data_graph <- with(data_new,
                   data.frame("Estimates" = c(h, tauE_est, tauA_est),
                              "Ic_Min" = c(IcMin_h, tauE_IcMin, tauA_IcMin),
                              "Ic_Max" = c(IcMax_h, tauE_IcMax, tauA_IcMax),
                              "Age" = rep(age, 3), "gender" = rep(gender, 3),
                              "Parameter" = rep(c("h","tauE","tauA"), each = 60) ))
levels(data_graph$Parameter) <- c("Heritability", "Genetic", "Environment")
ylim = list("1" = c(0.5,0.9), "1" = c(0.5, 0.9),
            "2" = c(6,12),"2" = c(6, 12),
            "3" = c(1, 7), "3" = c(1, 7))
xy1 <- xyplot(Estimates ~ Age | gender + Parameter, data = data_graph,
              ylim = ylim, type = "l", ly = data_graph$Ic_Min,
              xlab = c("Standardized age"),
              uy = data_graph$Ic_Max, scales = "free",
              #par.settings = ps,
              cty = "bands", alpha = 0.25, 
              prepanel = prepanel.cbH,
              panel = panel.cbH)
useOuterStrips(combineLimits(xy1,adjust.labels = TRUE))
```

<!---------------------------------------------------------------------- -->

[`mglm4twin`]: https://github.com/wbonat/mglm4twin
[mglm4twin/README]: https://github.com/wbonat/mglm4twin/blob/master/README.md
