---
title: "BPD and RDS on preterm infants: Bivariate dichotomous data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BPD and RDS on preterm infants: Bivariate dichotomous data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

The `mglm4twin` package implements multivariate generalized
linear models for twin and family data as proposed by Bonat and Hjelmborg (2020).
The core fit function `mglm4twin` is employed for fitting a set of models. 
In this introductory vignette we restrict ourselves to model 
bivariate binary outcomes. We present a set of special specifications of the standard ACE
for binary trait in the context of twin data. This vignette reproduces the data analysis presented in the Section 5.3 Dataset 3: Bivariate dichotomous data of Bonat and Hjelmborg (2020). 

## Install and loading the mglm4twin package

To install the stable version of [`mglm4twin`][], use
`devtools::install_git()`. For more information, visit [mglm4twin/README].

```{r, eval=FALSE}
library(devtools)
install_git("wbonat/mglm4twin")
```

```{r, eval=TRUE, error=FALSE, message=FALSE, warning=FALSE}
library(mglm4twin)
packageVersion("mglm4twin")
```

## Data set

In this example, we use the dataset analysed by Feng et al. (2009) regarding bronchopulmonary dysplasia (BPD) and respiratory distress syndrome (RDS) on preterm infants. 
Both diseases are lung related and expected to have a genetic component. 
The dataset consists of 200 twin-pairs being 137 DZ and 63 MZ. Additionally, we considered the
covariates: birth weight (BW), gestation age (GA) and gender (male and female).

```{r}
data(bpdrds)
head(bpdrds)
```


## Model specification

The model specification is done in three steps:

   1. Specifying the linear predictor (mean model).
   2. Specifying the matrix linear predictor (twin dependence structure).
   3. Specifying link and variance function suitable to response variable type.

### Specyfing the linear predictor

In the **bpdrds** data set we have three covariates available for composing the linear predictor. Thus, the linear predictor is given by

```{r}
form_BPD <- BPD ~ BW + GA + gender + Group*Twin_pair
form_RDS <- RDS ~ BW + GA + gender + Group*Twin_pair
```

It is important to note that in this example we have two outcomes. Consequently, we have to specify a linear predictor for each of them.

### Specyfing the matrix linear predictor

The matrix linear predictor describes the twin dependence structure. The `mglm4twin` package provides a set of pre-specified covariance structures. However, the users can easily specify their own covariance structure through a linear combination of known matrices. In this example, we are going to specify a set of special cases of the standard ACE models for twin data. Furthermore, for comparison purposes we are going to fit univariate and bivariate models. 

```{r}
## Univariate models
uni_E <- mt_twin(N_DZ = 137, N_MZ = 63, n_resp = 1, model = "E")
uni_AE <- mt_twin(N_DZ = 137, N_MZ = 63, n_resp = 1, model = "AE")
uni_CE <- mt_twin(N_DZ = 137, N_MZ = 63, n_resp = 1, model = "CE")
uni_ACE <- mt_twin(N_DZ = 137, N_MZ = 63, n_resp = 1, model = "ACE")

## Bivariate models
biv_E <- mt_twin(N_DZ = 137, N_MZ = 63, n_resp = 2, model = "E")
biv_AE <- mt_twin(N_DZ = 137, N_MZ = 63, n_resp = 2, model = "AE")
biv_CE <- mt_twin(N_DZ = 137, N_MZ = 63, n_resp = 2, model = "CE")
biv_ACE <- mt_twin(N_DZ = 137, N_MZ = 63, n_resp = 2, model = "ACE")
```

### Specyfing link and variance functions

The specificantion of the link and variance functions depends on the response variable type. In the case of the dichotomous data the `mglm4twin`package offers
the `logit`, `probit`, `cauchit`, `cloglog` and `loglog` link functions. For the variance function the `binomialP` represents the standard binomial variance function, i.e. $\mu(1-\mu)$ which is suitable for binary outcomes. In this example,
we follow Bonat and Hjelmborg (2020) and use the `logit` link function and the `binomialP` variance function for both outcomes.

```{r}
link = c("logit", "logit")
variance = c("binomialP", "binomialP")
```

It is interesting to note that the user should specify the link and variance functions for each outcome.

## Model fitting

The main fitting function in the `mglm4twin` package is the `mglm4twin()` function.
It is the same function for univariate and multivariate models. The main difference
is that now we should pass a vector of linear predictors, link and variance functions.
In this example, first we are going to fit the univariate models for each response variable and then the bivariate counterparts. 

```{r}
## Univariate fit

# Univariate E model
fitE_BPD <- mglm4twin(linear_pred = c(form_BPD), matrix_pred = uni_E,
                      link = c("logit"), variance = c("binomialP"),
                      data = bpdrds)
fitE_RDS <- mglm4twin(linear_pred = c(form_RDS), matrix_pred = uni_E,
                      link = c("logit"), variance = c("binomialP"),
                      data = bpdrds)

# Univariate AE model
fitAE_BPD <- mglm4twin(linear_pred = c(form_BPD), matrix_pred = uni_AE,
                      link = c("logit"), variance = c("binomialP"),
                      data = bpdrds)
fitAE_RDS <- mglm4twin(linear_pred = c(form_RDS), matrix_pred = uni_AE,
                       link = c("logit"), variance = c("binomialP"),
                       data = bpdrds)

# Univariate CE model
fitCE_BPD <- mglm4twin(linear_pred = c(form_BPD), matrix_pred = uni_CE,
                       link = c("logit"), variance = c("binomialP"),
                       data = bpdrds)
fitCE_RDS <- mglm4twin(linear_pred = c(form_RDS), matrix_pred = uni_CE,
                       link = c("logit"), variance = c("binomialP"),
                       data = bpdrds)

# Univariate ACE model
fitACE_BPD <- mglm4twin(linear_pred = c(form_BPD), matrix_pred = uni_ACE,
                       link = c("logit"), variance = c("binomialP"),
                       data = bpdrds)
fitACE_RDS <- mglm4twin(linear_pred = c(form_RDS), matrix_pred = uni_ACE,
                        link = c("logit"), variance = c("binomialP"),
                        data = bpdrds)

## Bivariate fit

# E model
fitE_biv <- mglm4twin(linear_pred = c(form_BPD, form_RDS), matrix_pred = biv_E, 
                      link = c("logit","logit"), 
                      variance = c("binomialP","binomialP"), 
                      data = bpdrds)
# AE model
fitAE_biv <- mglm4twin(linear_pred = c(form_BPD, form_RDS), matrix_pred = biv_AE, 
                       link = c("logit","logit"), 
                       variance = c("binomialP","binomialP"), 
                       data = bpdrds)
# CE model
fitCE_biv <- mglm4twin(linear_pred = c(form_BPD, form_RDS), matrix_pred = biv_CE, 
                       link = c("logit","logit"), 
                       variance = c("binomialP","binomialP"), 
                       data = bpdrds)
# ACE model
fitACE_biv <- mglm4twin(linear_pred = c(form_BPD, form_RDS), matrix_pred = biv_ACE, 
                        link = c("logit", "logit"),
                        variance = c("binomialP","binomialP"),
                        data = bpdrds)
```

## Model output

The `mglm4twin` package offers a general `summary()` function that can be customized to show the dispersion components (standard output) or measures of interest (`biometric = TRUE`) as genetic correlation and heritability. 

## Model comparison

For model comparison the `mglm4twin` package offers the `gof()`function for computing
pseudo version of the log-likelihood, AIC, BIC and pKLIC information criterio.
It is intereting to note that the `gof()`function allows to combine two univariate
fits for jointly computing the goodness-of-fit measures under the assumption of
zero correlation between the outcomes.

```{r}
## Univariate models
uni <- round(rbind("E" = gof(list(fitE_BPD, fitE_RDS)), 
                   "AE" = gof(list(fitAE_BPD, fitAE_RDS)) ,
                   "CE" = gof(list(fitCE_BPD, fitCE_RDS)),
                   "ACE" = gof(list(fitACE_BPD, fitACE_RDS))), 2)

## Bivariate models
multi <- round(rbind("E" = gof(fitE_biv), "AE" = gof(fitAE_biv), 
                     "CE" = gof(fitCE_biv), "ACE" = gof(fitACE_biv)), 2)

```

## Reproducing Table 3 of Bonat and Hjelmborg (2020)

```{r, echo = FALSE, results = "hide"}
Table3 <- cbind(t(uni)[c(1,3,5,2),c(4,3,2,1)], t(multi)[c(1,3,5,2),c(4,3,2,1)])
```

```{r}
Table3
```

## Reproducing Table 4 of Bonat and Hjelmborg (2020)

For reprocucing Table 4 of Bonat and Hjelmborg (2020), we have to compute the
genetic and environment correlation along with the heritability, environmentability
and bivariate heritability and environmentability. The `mglm4twin`offers some
facilities for such computation, see the code below.

```{r}
## Genetic correlation
tab_rho_a <- mt_compute_rho(Estimates = coef(fitAE_biv, model = "AE"), 
                            vcov = vcov(fitAE_biv, model = "AE"),
                            component = "A", n_resp = 2)

## Environment correlation
tab_rho_e <- mt_compute_rho(Estimates = coef(fitAE_biv, model = "AE"), 
                            vcov = vcov(fitAE_biv, model = "AE"),
                            component = "E", n_resp = 2)
```

The other measures can be extracted from the `summary()`.

```{r, echo = FALSE}
output <- summary(fitAE_biv, biometric = TRUE, model = "AE")
# Univariate heritability
h2 <- output$A_main$Estimates
h2_std <- output$A_main$std.error
# Bivariate heritability
h_biv <- output$A_cross$Estimates
h_biv_std <- output$A_cross$std.error
# Univariate environmentability
e2 <- output$Env_main$Estimates
e2_std <- output$Env_main$std.error
e_biv <- output$Env_cross$Estimates
e_biv_std <- output$Env_cross$std.error
HH <- c(h2[1],h2_std[1],h2[2],h2_std[2],h_biv, h_biv_std)
EE <- c(e2[1],e2_std[1],e2[2],e2_std[2],e_biv, e_biv_std)
G <- c(rep(NA,4), tab_rho_a$Estimates, tab_rho_a$std.error)
E <- c(rep(NA,4), tab_rho_e$Estimates, tab_rho_e$std.error)
Table4 <- round(rbind(HH,EE,G,E), 2)
colnames(Table4) <- c("BPD", "BPD_std", "RDS", "RDS_std", "BPD_RDS", "BPD_RDS_std")
rownames(Table4) <- c("Heritability", "Environmentability", "Genetic Cor", "Environment Cor")
```

```{r}
Table4
```

<!---------------------------------------------------------------------- -->

[`mglm4twin`]: http://github.com/wbonat/mglm4twin
[mglm4twin/README]: http://github.com/wbonat/mglm4twin/blob/master/README.md
