---
title: "Multivariate bounded data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multivariate bounded data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

The `mglm4twin` package implements multivariate generalized
linear models for twin and family data as proposed by Bonat and Hjelmborg (2020).
The core fit function `mglm4twin` is employed for fitting a set of models. 
In this introductory vignette we restrict ourselves to model 
multivariate bounded outcomes. 
We present a set of special specifications of the standard ACE
for multivariate bounded traits in the context of twin data. 
This vignette reproduces the data analysis presented in the Section 5.6 Dataset 6: Multivariate bounded data of Bonat and Hjelmborg (2020). 

## Install and loading the mglm4twin package

To install the stable version of [`mglm4twin`][], use
`devtools::install_git()`. For more information, visit [mglm4twin/README].

```{r, eval=FALSE}
library(devtools)
install_git("wbonat/mglm4twin")
```

```{r, eval=TRUE, error=FALSE, message=FALSE, warning=FALSE}
library(mglm4twin)
packageVersion("mglm4twin")
```

## Data set

In this example, we simulate a dataset as similar as possible to a real dataset
concernings three bounded outcomes reflecting mental conditions. The data set 
consists of 828 (524 DZ and 274 MZ) twin pairs. The outcomes are obtained based 
on questionaries and includes the Mini Mental State Examination (MMSE) taking values
on the interval $[0\ldots, 30]$. The second outcome corresponds to depression
symptomatology which was evaluated based on an adaptation of the depression 
section of the Cambridge Mental Disorders of the Elderly Examination (CAMDEX). 
The depression scale used here is a composite of responses to 17 depression items
and taking integer values in the interval $[0, \ldots, 30]$ and is highly 
left-skewed in distribution in any population-based sample. The third outcome is a Social Activity 
scale based on six items that assess the frequency with which the individual is 
engaged with others and mental pursuits. We opted to simulate such a data set
because we can show how to simulate the data and make it available for the users.
For simulating the data set, we are going to use the `SimCorMultRes` package.

```{r}
## Loading extra packages
require(Matrix)
require(SimCorMultRes)
require(mglm4twin)

## Setting model parameters
E <- c(0.75, 0.7, 0.65, -0.3, 0.25, -0.4)
A <- c(0.25,0.3, 0.35, -0.15, 0.20, -0.2)
C <- rep(0, 6)
tau = c(E, A, C)

## Twin structure
DZ = mt_twin(N_DZ = 1, N_MZ = 0, n_resp = 3, model = "ACE")
MZ = mt_twin(N_DZ = 0, N_MZ = 1, n_resp = 3, model = "ACE")
Omega_DZ <- as.matrix(mt_matrix_linear_predictor(tau = tau, Z = DZ))
Omega_MZ <- as.matrix(mt_matrix_linear_predictor(tau = tau, Z = MZ))

## Regression structures
sex <- c(rep("Male", 220/2), rep("Female", 334/2), 
         rep("Male", 116/2), rep("Female", 158/2))
zyg <- c(rep("DZ", 554/2), rep("MZ", 274/2))
set.seed(123)
Age <- rbeta(414, shape1 = 0.3*2, shape2 = 0.7*2)*20 + 70
age_std <- (Age - mean(Age))/var(Age)
X <- model.matrix(~ sex + age_std)
beta1 <- c(1.6956, 0.0584, -0.2576)
mu1 <- exp(X%*%beta1)/(1+exp(X%*%beta1))
beta2 <- c(-1.7930, 0.0875, 0.2382)
mu2 <- exp(X%*%beta2)/(1+exp(X%*%beta2))
beta3 <- c(-0.5363, -0.05138, -0.1528)
mu3 <- exp(X%*%beta3)/(1+exp(X%*%beta3))

## Marginal distributions
dist <- c("qbeta","qbeta","qbeta")
invcdfnames <- rep(dist, each = 2)

## Simulating data set
Y1_MZ <- list()
Y2_MZ <- list()
Y3_MZ <- list()

Y1_DZ <- list()
Y2_DZ <- list()
Y3_DZ <- list()
set.seed(181185)
phi = 5
for(i in 1:137) {
  paramslists <- list(
    m1 = list(shape1 = mu1[i]*phi, shape2 = (1-mu1[i])*phi),
    m1 = list(shape1 = mu1[i]*phi, shape2 = (1-mu1[i])*phi),
    m2 = list(shape1 = mu2[i]*phi, shape2 = (1-mu2[i])*phi),
    m2 = list(shape1 = mu2[i]*phi, shape2 = (1-mu2[i])*phi),
    m3 = list(shape1 = mu3[i]*phi, shape2 = (1-mu3[i])*phi),
    m3 = list(shape1 = mu3[i]*phi, shape2 = (1-mu3[i])*phi))
  Y_MZ <- rnorta(R = 1, cor.matrix = Omega_MZ,
                 distr = invcdfnames, qparameters = paramslists)
  Y1_MZ[[i]] <- Y_MZ[1:2]
  Y2_MZ[[i]] <- Y_MZ[3:4]
  Y3_MZ[[i]] <- Y_MZ[5:6]
}
set.seed(181185)
for(i in 138:414) {
  paramslists <- list(
    m1 = list(shape1 = mu1[i]*phi, shape2 = (1-mu1[i])*phi),
    m1 = list(shape1 = mu1[i]*phi, shape2 = (1-mu1[i])*phi),
    m2 = list(shape1 = mu2[i]*phi, shape2 = (1-mu2[i])*phi),
    m2 = list(shape1 = mu2[i]*phi, shape2 = (1-mu2[i])*phi),
    m3 = list(shape1 = mu3[i]*phi, shape2 = (1-mu3[i])*phi),
    m3 = list(shape1 = mu3[i]*phi, shape2 = (1-mu3[i])*phi))
  Y_DZ <- rnorta(R = 1, cor.matrix = Omega_DZ,
                 distr = invcdfnames, qparameters = paramslists)
  Y1_DZ[[i]] <- Y_DZ[1:2]
  Y2_DZ[[i]] <- Y_DZ[3:4]
  Y3_DZ[[i]] <- Y_DZ[5:6]
}

Y1 <- c(do.call(c, Y1_DZ), do.call(c, Y1_MZ))
Y2 <- c(do.call(c, Y2_DZ), do.call(c, Y2_MZ))
Y3 <- c(do.call(c, Y3_DZ), do.call(c, Y3_MZ))

data <- data.frame("Y1" = Y1, "Y2" = Y2, "Y3" = Y3, "twin_id" = rep(1:2, 414),
                   "zyg" = rep(zyg, each = 2), "sex" = rep(sex, each = 2), 
                   "age_std" = rep(age_std, each = 2))
```


## Model specification

The model specification is done in three steps:

   1. Specifying the linear predictor (mean model).
   2. Specifying the matrix linear predictor (twin dependence struture).
   3. Specifying link and variance function suitable to be response variable type.

### Specyfing the linear predictor

In the example we have the covariates sex and age for composing the linear predictor. 
Thus, the linear predictor is given by

```{r}
form_Y1 <- c(Y1 ~ sex + age_std)
form_Y2 <- c(Y2 ~ sex + age_std)
form_Y3 <- c(Y3 ~ sex + age_std)
```

### Specyfing the matrix linear predictor

The matrix linear predictor describes the twin dependence structure. The `mglm4twin` package provides a set of pre-specified covariance structures. However, the users can easily specify thein own covariance structure through a linear combination of known matrices. In this example, we are going to specify a set of special cases of the standard ACE models for twin data.

```{r}
ACE = mt_twin(N_DZ = 554/2, N_MZ = 274/2, n_resp = 3, model = "ACE")
AE = mt_twin(N_DZ = 554/2, N_MZ = 274/2, n_resp = 3, model = "AE")
CE = mt_twin(N_DZ = 554/2, N_MZ = 274/2, n_resp = 3, model = "CE")
E = mt_twin(N_DZ = 554/2, N_MZ = 274/2, n_resp = 3, model = "E")
```

The function `mt_twin()` has four main arguments: `N_DZ` and `N_MZ` the number of dizygotic and monozygotic twin, respectively. `n_resp` number of response variables, in this case just $1$ and the `model` a string given the model's name.
It is **extreme important** to note that the data set should be ordered in a way that all DZ twin appear first and then all MZ twin and the twin pair should appear together, i.e. twin 2 after twin 1. 

### Specyfing link and variance functions

The specificantion of the link and variance functions depends on the response variable type. In the case of the bounded data the `mglm4twin`package offers
the `logit`, `probit`, `cauchit`, `cloglog` and `loglog` link functions. 
For the variance function the `binomialP` represents the standard binomial variance function, i.e. $\mu(1-\mu)$ which is suitable for bounded outcomes. In this example,
we follow Bonat and Hjelmborg (2020) and use the `logit` link function and the `binomialP` variance function.

```{r}
link = rep("logit", 3)
variance = rep("binomialP", 3)
```

## Model fitting

The main fitting function in the `mglm4twin` package is the `mglm4twin()` function.
it was designed to be similar to standard model fitting functions in `R` such as `lm()` and `glm()`. 

```{r, warning = FALSE}
fit_ACE <- mglm4twin(linear_pred = c(form_Y1, form_Y2, form_Y3), 
                     matrix_pred = ACE, link = link,
                     variance = variance, 
                     data = data)
fit_AE <- mglm4twin(linear_pred = c(form_Y1, form_Y2, form_Y3), 
                    matrix_pred = AE, link = link,
                    variance = variance, 
                    data = data)
fit_CE <- mglm4twin(linear_pred = c(form_Y1, form_Y2, form_Y3), 
                    matrix_pred = CE, link = link,
                    variance = variance, 
                    data = data)
fit_E <- mglm4twin(linear_pred = c(form_Y1, form_Y2, form_Y3), 
                    matrix_pred = E, link = link,
                    variance = variance, 
                    data = data)
```

## Reproducing Table 8 of Bonat and Hjelmborg (2020)

```{r}
rbind(gof(fit_E), gof(fit_AE), gof(fit_CE), gof(fit_ACE))[,c(1,2,3,5)]
```

## Reproducing Table 9 of Bonat and Hjelmborg (2020)

Table 9 is done by collecting the results from the model's summary.

```{r}
summary(fit_AE, model = "AE", biometric = TRUE)
```



<!---------------------------------------------------------------------- -->

[`mglm4twin`]: http://github.com/wbonat/mglm4twin
[mglm4twin/README]: http://github.com/wbonat/mglm4twin/blob/master/README.md
