---
title: "Psychiatric disorder: Dichotomous data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Psychiatric disorder: Dichotomous data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

The `mglm4twin` package implements multivariate generalized
linear models for twin and family data as proposed by Bonat and Hjelmborg (2020).
The core fit function `mglm4twin` is employed for fitting a set of models. 
In this introductory vignette we restrict ourselves to model a
binary trait. We present a set of special specifications of the standard ACE
for binary traits in the context of twin data. 
This vignette reproduces the data analysis presented in the Section 5.1 
Dataset 1: Dichotomous data of Bonat and Hjelmborg (2020). 

## Install and loading the mglm4twin package

To install the stable version of [`mglm4twin`][], use
`devtools::install_git()`. For more information, visit [mglm4twin/README].

```{r, eval=FALSE}
library(devtools)
install_git("wbonat/mglm4twin")
```

```{r, eval=TRUE, error=FALSE, message=FALSE, warning=FALSE}
library(mglm4twin)
packageVersion("mglm4twin")
```

## Data set

The dataset concerns psychiatric disorders in 1030 (440 DZ and 590 MZ) Caucasian 
female twin-pairs sampled from the Virginia Twin Registry. Lifetime psychiatric 
illness is a binary trait and was diagnosed using an adapted version of the 
Structured Clinical Interview for DSM-II-R Diagnosis. The dataset was analysed by 
Neale and Maes (2004) and Rabe-Hesketh et al. (2008) through generalized linear 
mixed models. The main goal of this study is to measure the genetic influence on 
the binary trait.


```{r}
data(psydis)
head(psydis)
```


## Model specification

The model specification is done in three steps:

   1. Specifying the linear predictor (mean model).
   2. Specifying the matrix linear predictor (twin dependence struture).
   3. Specifying link and variance function suitable to be response variable type.

### Specyfing the linear predictor

In the **psydis** data set we have no covariate for composing the linear predictor.
Thus, the linear predictors are given by

```{r}
linear_pred <- y ~ 1
```

### Specyfing the matrix linear predictor

The matrix linear predictor describes the twin dependence structure. The `mglm4twin` package provides a set of pre-specified covariance structures. However, the users can easily specify thein own covariance structure through a linear combination of known matrices. In this example, we are going to specify a set of special cases of the standard ACE models for twin data.

```{r}
ex1_E <- mt_twin(N_DZ = 440, N_MZ = 590, n_resp = 1, model = "E")
ex1_AE <- mt_twin(N_DZ = 440, N_MZ = 590, n_resp = 1, model = "AE")
ex1_CE <- mt_twin(N_DZ = 440, N_MZ = 590, n_resp = 1, model = "CE")
ex1_ACE <- mt_twin(N_DZ = 440, N_MZ = 590, n_resp = 1, model = "ACE")
```

The function `mt_twin()` has four main arguments: `N_DZ` and `N_MZ` the number of dizygotic and monozygotic twin, respectively. `n_resp` number of response variables, in this case just $1$ and the `model` a string given the model's name.
It is **extreme important** to note that the data set should be ordered in a way that all DZ twin appear first and then all MZ twin and the twin pair should appear together, i.e. twin 2 after twin 1. 

### Specyfing link and variance functions

The specificantion of the link and variance functions depends on the response variable type. In the case of the dichotomous data the `mglm4twin`package offers
the `logit`, `probit`, `cauchit`, `cloglog` and `loglog` link functions. For the variance function the `binomialP` represents the standard binomial variance function, i.e. $\mu(1-\mu)$ which is suitable for binary outcomes. In this example,
we follow Bonat and Hjelmborg (2020) and use the `logit` link function and the `binomialP` variance function.

```{r}
link = "logit"
variance = "binomialP"
```

## Model fitting

The main fitting function in the `mglm4twin` package is the `mglm4twin()` function.
it was designed to be similar to standard model fitting functions in `R` such as `lm()` and `glm()`. 

```{r}
fit_ex1_E <- mglm4twin(linear_pred = c(linear_pred), matrix_pred = ex1_E, 
                       link = c(link), variance = c(variance), data = psydis)
fit_ex1_AE <- mglm4twin(linear_pred = c(linear_pred), matrix_pred = ex1_AE, 
                        link = c(link), variance = c(variance), data = psydis)
fit_ex1_CE <- mglm4twin(linear_pred = c(linear_pred), matrix_pred = ex1_CE, 
                        link = c(link), variance = c(variance), data = psydis)
fit_ex1_ACE <- mglm4twin(linear_pred = c(linear_pred), matrix_pred = ex1_ACE, 
                         link = c(link), variance = c(variance), data = psydis)
```

## Model output

The `mglm4twin` package offers a general `summary()` function that can be customized to show the dispersion components (standard output) or measures of interest (`biometric = TRUE`) as genetic correlation and heritability. 

#### Standard output

```{r}
summary(fit_ex1_ACE, model = "ACE") ## Standard output
```


#### Biometric output

```{r}
summary(fit_ex1_ACE, model = "ACE", biometric = TRUE) ## Standard output
```

## Reproducing Table 1 of Bonat and Hjelmborg (2020)

```{r, echo = FALSE, results = "hide"}
Estimates <- cbind(c(coef(fit_ex1_E, model = "E")$Estimates[2], rep(NA, 2)),
c(coef(fit_ex1_AE, model = "AE")$Estimates[2:3], rep(NA, 1)),
c(coef(fit_ex1_CE, model = "CE")$Estimates[2], 0, coef(fit_ex1_CE, model = "CE")$Estimates[3]),
coef(fit_ex1_ACE, model = "ACE")$Estimates[2:4])

Std_error <- cbind(c(coef(fit_ex1_E, model = "E", 
                          std.error = TRUE)$std.error[2], rep(NA, 2)),
                   c(coef(fit_ex1_AE, model = "AE", 
                          std.error = TRUE)$std.error[2:3], rep(NA, 1)),
                   c(coef(fit_ex1_CE, model = "CE", 
                          std.error = TRUE)$std.error[2], 0, 
                     coef(fit_ex1_CE, model = "CE", 
                          std.error = TRUE)$std.error[3]),
                   coef(fit_ex1_ACE, model = "ACE", 
                        std.error = TRUE)$std.error[2:4])
ll <- c(as.numeric(plogLik(fit_ex1_E, verbose = FALSE)), 
        as.numeric(plogLik(fit_ex1_AE, verbose = FALSE)),
        as.numeric(plogLik(fit_ex1_CE, verbose = FALSE)),
        as.numeric(plogLik(fit_ex1_ACE, verbose = FALSE)))
Table1 <- cbind(Estimates[,1], Std_error[,1], Estimates[,2], Std_error[,2],
                Estimates[,3], Std_error[,3], Estimates[,4], Std_error[,4])
h2 <- summary(fit_ex1_AE, biometric = TRUE, model = "AE")
h2_ACE <- summary(fit_ex1_ACE, biometric = TRUE, model = "ACE")
H <- c(NA, NA, c(h2$A_main$Estimates, h2$A_main$std.error), NA, NA,
       c(h2_ACE$A_main$Estimates, h2_ACE$A_main$std.error))
Table1 <- round(rbind(Table1, H, ll), 2)
rownames(Table1) <- c("E", "A", "C", "h2", "ll")
colnames(Table1) <- c("Estimate", "Std_Error", "Estimate", "Std_Error", 
                      "Estimate", "Std_Error", "Estimate", "Std_Error")
```

```{r}
Table1
```

<!---------------------------------------------------------------------- -->

[`mglm4twin`]: http://github.com/wbonat/mglm4twin
[mglm4twin/README]: http://github.com/wbonat/mglm4twin/blob/master/README.md
