---
title: "Sleep’s quality: Mixed types of outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sleep’s quality: Mixed types of outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

The `mglm4twin` package implements multivariate generalized
linear models for twin and family data as proposed by Bonat and Hjelmborg (2020).
The core fit function `mglm4twin` is employed for fitting a set of models. 
In this introductory vignette we restrict ourselves to model 
mixed types of outcomes. We present a set of special specifications of the standard ACE
for the mixed of continuous and bounded traits in the context of twin data. 
This vignette reproduces the data analysis presented in the Section 5.5 
Dataset 5: Mixed types of traits data of Bonat and Hjelmborg (2020). 

## Install and loading the mglm4twin package

To install the stable version of [`mglm4twin`][], use
`devtools::install_git()`. For more information, visit [mglm4twin/README].

```{r, eval=FALSE}
library(devtools)
install_git("wbonat/mglm4twin")
```

```{r, eval=TRUE, error=FALSE, message=FALSE, warning=FALSE}
library(mglm4twin)
packageVersion("mglm4twin")
```

## Data set

The data set concerning sleep’s quality in a sample of 250 (135 DZ and 116 MZ) 
Danish twin pairs. The traits are cortisone levels when waking up (T0) and PSQI 
(Pittsburgh Sleep Quality Index). The first one is a continuous trait, while the 
second one is a scale varying from 0 to 21 (slower values better sleep’s quality).
The goal of this example is to show how to deal with mixed types of traits in the 
context of twin data. It is a simulated data set based on the parameter estimates
of the real data fit. 

```{r}
data(t0psqi)
head(t0psqi)
```


## Model specification

The model specification is done in three steps:

   1. Specifying the linear predictor (mean model).
   2. Specifying the matrix linear predictor (twin dependence struture).
   3. Specifying link and variance function suitable to be response variable type.

### Specyfing the linear predictor

In the **t0psqi** data set we have the covariates `Age`, `Type`, `Gender` and 
`Group` for composing the linear predictor for both outcomes. 
Thus, the linear predictors are given by

```{r}
form_T0 <- T0 ~ Age + Gender + Group + Type*Twin_pair
form_PSQI <- PSQI ~ Age + Gender + Group + Type*Twin_pair
```

### Specyfing the matrix linear predictor

The matrix linear predictor describes the twin dependence structure. The `mglm4twin` package provides a set of pre-specified covariance structures. However, the users can easily specify thein own covariance structure through a linear combination of known matrices. In this example, we are going to specify a set of special cases of the standard ACE models for twin data.

```{r}
E <- mt_twin(N_DZ = 135, N_MZ = 116, n_resp = 2, model = "E")
AE <- mt_twin(N_DZ = 135, N_MZ = 116, n_resp = 2, model = "AE")
CE <- mt_twin(N_DZ = 135, N_MZ = 116, n_resp = 2, model = "CE")
ACE <- mt_twin(N_DZ = 135, N_MZ = 116, n_resp = 2, model = "ACE")
```

The function `mt_twin()` has four main arguments: `N_DZ` and `N_MZ` the number 
of dizygotic and monozygotic twin, respectively. `n_resp` number of response 
variables, in this case $1$ and the `model` a string given the model's name.
It is **extreme important** to note that the data set should be ordered in a way 
that all DZ twin appear first and then all MZ twin and the twin pair should 
appear together, i.e. twin 2 after twin 1. 

## Model fitting

The main fitting function in the `mglm4twin` package is the `mglm4twin()` function.
it was designed to be similar to standard model fitting functions in `R` such as `lm()` and `glm()`. In this example, we are going to fit Gaussian bivariate models and
a more felxible model where we use a Tweedie model for the T0 trait and
a flexible quasi-beta model for the PSQI index. We opted to do this, because
Gaussian models are a frequently choice for this types of data.



```{r}
## Gaussian model (link = 'identity' and variance = 'constant')
Gauss_E <- mglm4twin(linear_pred = c(form_T0,  form_PSQI), 
                     matrix_pred = E, data = t0psqi)
Gauss_AE <- mglm4twin(linear_pred = c(form_T0,  form_PSQI), 
                      matrix_pred = AE, data = t0psqi)
Gauss_CE <- mglm4twin(linear_pred = c(form_T0,  form_PSQI), 
                      matrix_pred = CE, data = t0psqi)
Gauss_ACE <- mglm4twin(linear_pred = c(form_T0,  form_PSQI), 
                       matrix_pred = ACE, data = t0psqi)

## Mixed types
fit_E <- mglm4twin(linear_pred = c(form_T0,  form_PSQI), 
                   matrix_pred = E, 
                   link = c("log","logit"), 
                   variance = c("tweedie","binomialP"), 
                   control_algorithm = list(tuning = 0.25, max_iter = 100),
                   power_fixed = c(F,F), data = t0psqi)
fit_AE <- mglm4twin(linear_pred = c(form_T0,  form_PSQI), 
                    matrix_pred = AE, 
                    link = c("log","logit"), 
                    variance = c("tweedie","binomialP"), 
                    control_algorithm = list(tuning = 0.25, max_iter = 100),
                    power_fixed = c(F,F), data = t0psqi)
fit_CE <- mglm4twin(linear_pred = c(form_T0,  form_PSQI), 
                    matrix_pred = CE, 
                    link = c("log","logit"), 
                    variance = c("tweedie","binomialP"), 
                    control_algorithm = list(tuning = 0.25, max_iter = 100),
                    power_fixed = c(F,F), data = t0psqi)
fit_ACE <- mglm4twin(linear_pred = c(form_T0,  form_PSQI), 
                     matrix_pred = ACE, 
                     link = c("log","logit"), 
                     variance = c("tweedie","binomialP"), 
                     control_algorithm = list(tuning = 0.25, max_iter = 100),
                     power_fixed = c(F,F), data = t0psqi)
```

## Reproducing Table 6 of Bonat and Hjelmborg (2020)

```{r, echo = FALSE, results = "hide"}
Table6 <- cbind(rbind(gof(Gauss_E), gof(Gauss_AE), gof(Gauss_CE), 
                      gof(Gauss_ACE))[,-c(4)],
                rbind(gof(fit_E), gof(fit_AE), gof(fit_CE), 
                      gof(fit_ACE))[,-c(4)])

```

```{r}
Table6
```

## Reproducing Table 7 of Bonat and Hjelmborg (2020)

Table 7 is easily done by collecting the values from the model's summary.

```{r}
summary(fit_AE, model = "AE", biometric = TRUE)
```

<!---------------------------------------------------------------------- -->

[`mglm4twin`]: http://github.com/wbonat/mglm4twin
[mglm4twin/README]: http://github.com/wbonat/mglm4twin/blob/master/README.md
