---
title: "Anthropometric measures: Bivariate continuous data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anthropometric measures: Bivariate continuous data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Abstract

The `mglm4twin` package implements multivariate generalized
linear models for twin and family data as proposed by Bonat and Hjelmborg (2020).
The core fit function `mglm4twin` is employed for fitting a set of models. 
In this introductory vignette we restrict ourselves to model 
bivariate continuous outcomes. We present a set of special specifications of the standard ACE
for continuous trait in the context of twin data. This vignette reproduces the data analysis presented in the Section 5.4 Dataset 4: Bivariate continuous data of Bonat and Hjelmborg (2020). 

## Install and loading the mglm4twin package

To install the stable version of [`mglm4twin`][], use
`devtools::install_git()`. For more information, visit [mglm4twin/README].

```{r, eval=FALSE}
library(devtools)
install_git("wbonat/mglm4twin")
```

```{r, eval=TRUE, error=FALSE, message=FALSE, warning=FALSE}
library(mglm4twin)
packageVersion("mglm4twin")
```

## Data set

The dataset `anthro` contains anthropometric measures (weight and height) on 861 (327 DZ and 534 MZ) twin-pairs. The data set is available as an example in the OpenMx package (Neale et al., 2016).
The main goal of this example is to illustrate how to deal with a fairly common case of bivariate continuous traits in the context of twin studies. Furthermore, we explore the flexibility of our proposed model class and model the dispersion components as a linear combination of a covariate of interest.

```{r}
data(anthro)
head(anthro)

## Standardize
anthro$age <- (anthro$age - mean(anthro$age))/sd(anthro$age)
anthro$weight <- (anthro$weight - mean(anthro$weight))/sd(anthro$weight)
anthro$height <- (anthro$height - mean(anthro$height))/sd(anthro$height)

```

## Model specification

The model specification is done in three steps:

   1. Specifying the linear predictor (mean model).
   2. Specifying the matrix linear predictor (twin dependence struture).
   3. Specifying link and variance function suitable to be response variable type.

### Specyfing the linear predictor

In the **anthro** data set we have three covariates available for composing the linear predictor. Thus, the linear predictor is given by

```{r}
form_Wt <- weight ~ age + Group*Twin_pair
form_Ht <- height ~ age + Group*Twin_pair
```

### Specyfing the matrix linear predictor

In this example, we are going to specify and fit only the two models presented
in Table 5 of Bonat and Hjelmborg (2020). The first model is an specification of
the ACE model where all dispersion components are modelled as a function of the
covariate age. The second is a special case of the former whre all non-significant
terms were excluded.

```{r}
## ACE model
biv0 <- list("formE1" = ~ age, "formE2" = ~ age, "formE12" = ~ age,
             "formA1" = ~ age, "formA2" = ~ age, "formA12" = ~ age,
             "formC1" = ~ age, "formC2" = ~ age, "formC12" = ~ age)
Z_biv0 <- mt_twin(N_DZ = 327, N_MZ = 534, n_resp = 2, model = "ACE",
                    formula = biv0, data = anthro)

## Special case of ACE model
biv4 <- list("formE1" = ~ age, "formE2" = ~ age, "formE12" = ~ 1,
             "formA1" = ~ age, "formA2" = ~ 1, "formA12" = ~ 1)
Z_biv4 <- mt_twin(N_DZ = 327, N_MZ = 534, n_resp = 2, model = "AE",
                  formula = biv4, data = anthro)
```

## Model fitting

The main fitting function in the `mglm4twin` package is the `mglm4twin()` function.
it was designed to be similar to standard model fitting functions in `R` such as `lm()` and `glm()`. 

```{r}
## Initial values
control_initial <- list()
control_initial$regression <- list("R1" = c(0.13, 0.10, -0.20, -0.02, 0.037),
                                   "R2" = c(0.23, 0.01, -0.27, -0.11, 0.11))
control_initial$power <- list(c(0), c(0))
control_initial$tau <- c(0.15, 0, 0.12, rep(0,15))

fit_0 <- mglm4twin(linear_pred = c(form_Wt, form_Ht), matrix_pred = Z_biv0,
                          control_initial = control_initial,
                          control_algorithm = list(tuning = 0.5),
                          power_fixed = c(TRUE, TRUE),
                          data = anthro)

control_initial$tau <- c(0.15, 0, 0.12, rep(0,6))
fit_4 <- mglm4twin(linear_pred = c(form_Wt, form_Ht), matrix_pred = Z_biv4,
                   control_initial = control_initial,
                   control_algorithm = list(tuning = 0.5),
                   power_fixed = c(TRUE, TRUE),
                   data = anthro)
```

## Model output

The `mglm4twin` package offers a general `summary()` function that can be customized to show the dispersion components (standard output) or measures of interest (`biometric = TRUE`) as genetic correlation and heritability. 

#### Parameter estimates and standard errors

```{r, warning=FALSE, message=FALSE}
est0 <- aux_summary(fit_0, formula = biv0, type = "otimist", data = anthro)[,c(1,2,3,4)]
est4 <- aux_summary(fit_4, formula = biv4, type = "otimist", data = anthro)[,c(1,2,3,4)]
Table5 <- data.frame("Estimates" = est0$Estimates, 
                     "Std.error" = est0$Std..Error, 
                     "Z.value" = est0$Z.value,
                     "Estimates" = c(est4$Estimates[1:5], NA, est4$Estimates[6:8], 
                                     NA,est4$Estimates[9], rep(NA, 7)),
                     "Std.error" = c(est4$Std..Error[1:5], NA, est4$Std..Error[6:8], NA,  
                                     est4$Std..Error[9], rep(NA, 7)), 
                     "Z.value" = c(est4$Z.value[1:5], NA, est4$Z.value[6:8], NA, 
                                   est4$Z.value[9], rep(NA, 7)))
rownames(Table5) <- c("E1", "E1.age", "E2", "E2.age", "E12", "E12.age","A1","A1.age",
                      "A2", "A2.age", "A12", "A12.age", "C1", "C1.age", "C2", "C2.age",
                      "C12", "C12.age")
Table5 <- round(Table5, 2)
```

## Reproducing Table 5 of Bonat and Hjelmborg (2020)

```{r}
Table5
```

## Reproducing Figure 6 of Bonat and Hjelmborg (2020)

Computing measures of interest, heritability, genetic and phenotipic correlations.

```{r}
## Computing measures of interest
Age = seq(-1.90, 1.70, l = 100)
point <- fit_4$Covariance

# Heritability
h_wt <- (point[6] + point[7]*Age)/(point[1] + point[2]*Age + point[6] + point[7]*Age)
h_ht <- (point[8])/(point[3] + point[4]*Age + point[8])
h_ht_wt <- point[9]/(point[5] + point[9])

# Genetic and phenotipic correlation
r_G <- point[9]/(sqrt(point[6] + point[7]*Age)*sqrt(point[8]))
std_1 <- sqrt(point[1] + point[2]*Age + point[6] + point[7]*Age)
std_2 <- sqrt(point[3] + point[4]*Age + point[8])
r_P <- (point[9] + point[5])/(std_1*std_2)
```

Inference for complex models like the one in this example is quite challenging.
Thus, we opted to perform inference based on simulation from the assymptotic
distribution of the estimating function estimators.

```{r}
require(mvtnorm)
Age = seq(-1.90, 1.70, l = 50)

# Point estimates
point1 <- fit_4$Covariance

# Covariance matrix (otimist - Gaussian case -> maximum likelihood estimator)
COV <- -2*fit_4$joint_inv_sensitivity[11:19, 11:19]

## Simulating from the assymptotic ditribution
point <- rmvnorm(n = 10000, mean = point1, sigma = as.matrix(COV))
res_h_wt <- matrix(NA, nrow = 50, ncol = 10000)
res_h_ht <- matrix(NA, nrow = 50, ncol = 10000)
res_ht_wt <- matrix(NA, nrow = 50, ncol = 10000)
res_r_G <- matrix(NA, nrow = 50, ncol = 10000)
res_r_P <- matrix(NA, nrow = 50, ncol = 10000)
res_r_E <- matrix(NA, nrow = 50, ncol = 10000)

for(i in 1:10000) {
  h_wt <- (point[i,6] + point[i,7]*Age)/(point[i,1] + point[i,2]*Age + point[i,6] + point[i,7]*Age)
  res_h_wt[,i] <- h_wt
  h_ht <- (point[i,8])/(point[i,3] + point[i,4]*Age + point[i,8])
  res_h_ht[,i] <- h_wt
  h_ht_wt <- point[i,9]/(point[i,5] + point[i,9])
  res_ht_wt[,i] <- h_ht_wt
  r_G <- point[i,9]/(sqrt(point[i,6] + point[i,7]*Age)*sqrt(point[i,8]))
  res_r_G[,i] <- r_G
  std_1 <- sqrt(point[i,1] + point[i,2]*Age + point[i,6] + point[i,7]*Age)
  std_2 <- sqrt(point[i,3] + point[i,4]*Age + point[i,8])
  r_P <- (point[9] + point[5])/(std_1*std_2)
  res_r_P[,i] <- r_P
  r_E <- point[i,5]/(sqrt(point[i,1] + point[i,2]*Age)*sqrt(point[i,3] + point[i,4]*Age))
  res_r_E[,i] <- r_E
}
```

Based on the simulated values is easy to get point estimates and confidence intervals
for all measures of interest.

```{r}
Estimates <- c(rowMeans(res_h_wt), rowMeans(res_h_ht), 
               rowMeans(res_ht_wt),
               rowMeans(res_r_G), rowMeans(res_r_E),
               rowMeans(res_r_P))
Ic_Min <- c(apply(res_h_wt, 1, quantile, 0.025),
            apply(res_h_ht, 1, quantile, 0.025),
            apply(res_ht_wt, 1, quantile, 0.025),
            apply(res_r_G, 1, quantile, 0.025),
            apply(res_r_E, 1, quantile, 0.025),
            apply(res_r_P, 1, quantile, 0.025))
Ic_Max <- c(apply(res_h_wt, 1, quantile, 0.975),
            apply(res_h_ht, 1, quantile, 0.975),
            apply(res_ht_wt, 1, quantile, 0.975),
            apply(res_r_G, 1, quantile, 0.975),
            apply(res_r_E, 1, quantile, 0.975),
            apply(res_r_P, 1, quantile, 0.975))
data_graph <- data.frame("Estimates" = Estimates, "Ic_Min" = Ic_Min,
                         "Ic_Max" = Ic_Max, "Age" = rep(Age, 6), 
                         "Parameter" = rep(c("h1","h2","h12","G","E","P"), each = 50))
```

Probably, the most effective way to show such a complex results is through a
plot.

```{r}
require(lattice)
require(latticeExtra)
source("https://raw.githubusercontent.com/walmes/wzRfun/master/R/panel.cbH.R")
data_graph$Parameter <- factor(data_graph$Parameter, 
                               levels = c("h1","h2","h12","G","E","P"))
levels(data_graph$Parameter) <- c("Weight heritability", 
                                  "Height heritability", 
                                  "Biv. heritability",
                                  "Genetic correlation",
                                  "Environmental correlation",
                                  "Phenotypic correlation")

ylim = list("1" = c(0.75,1), "1" = c(0.75, 1),
            "3" = c(0.75,1), "2" = c(0.1, 1),
            "3" = c(0.1,1), "3" = c(0.1, 1))

xy1 <- xyplot(Estimates ~ Age | Parameter, data = data_graph, 
              ylim = ylim, 
              type = "l", ly = data_graph$Ic_Min, 
              xlab = c("Standardized age"),
              uy = data_graph$Ic_Max, scales = "free",
              cty = "bands", alpha = 0.25, prepanel = prepanel.cbH, 
              panel = panel.cbH)
```

```{r, echo = FALSE, out.width='99%', message=FALSE, fig.width=10, fig.height=6, fig.align='center', fig.cap = 'Genetic, environment and phenotypic correlations (first line). Body weight heritability, height heritability and bivariate heritability (second line) - Bivariate continuous Twin data.'}
xy1
```

<!---------------------------------------------------------------------- -->

[`mglm4twin`]: http://github.com/wbonat/mglm4twin
[mglm4twin/README]: http://github.com/wbonat/mglm4twin/blob/master/README.md
